---
title: Country Centroids
author: John Waller
date: '2019-03-19'
slug: country-centroids
categories:
  - GBIF
tags:
  - centroids
  - data quality
lastmod: '2019-02-21T10:23:10+01:00'
draft: yes
keywords: []
description: ''
authors: ''
comment: no
toc: ''
autoCollapseToc: no
postMetaInFooter: no
hiddenFromHomePage: no
contentCopyright: no
reward: no
mathjax: no
mathjaxEnableSingleDollar: no
mathjaxEnableAutoNumber: no
hideHeaderAndFooter: no
flowchartDiagrams:
  enable: no
  options: ''
sequenceDiagrams:
  enable: no
  options: ''
---

<img src = "/post/2019-02-21-country-centroids_files/japanCentroid.gif" style="display: block;margin-left: auto;margin-right: auto;"/>


**Country Centroids** are a known **data quality** issue within the [GBIF network](https://www.gbif.org/dataset/search). 

> Sometimes data publishers will not know the exact lat-lon location of a record and will enter the lat-long **center of the country** instead. This is a data  issue because **users might be unaware that an observation is pinned to a country center** and **assume** it is a **precise location**. 

Below I plot the top country centroids found on GBIF within at least 1km. Only the top unique point from each centroid is plotted, so there could be more occurrences near the centroids. 

```{r, echo=FALSE, warning=FALSE,message=FALSE,fig.width=8,fig.height=4}
library(leaflet)
widget = readRDS("C:/Users/ftw712/Desktop/country centroids/rda/leafletMapTopCentroidCountries.rda")
widget
```


A few countries have very many points falling on their country centroids. For example, **Brazil** and **Mexico** each have over **25 k** occurrence records within 1 km of their country centroid. 

<img src = "/post/2019-02-21-country-centroids_files/barPlotCountryCentroids.jpg" style="display: block;margin-left: auto;margin-right: auto;"/>


# Other regional or province centroids

Publishers might not only pin occurrences to the center of country but might also give the lat-lon centers of smaller adminstrative zones. For example the **United States does not have an obvious country centroid** but **Hawaii** has 704 occurrences near the **state-centroid**.   

<img src = "/post/2019-02-21-country-centroids_files/hawaiiCentroid.png" style="display: block;margin-left: auto;margin-right: auto;"/>


# Some countries like Monoco are too small to filter

Some countries and provinces are small enough that it does not make sense to filter them out even if they happen to be close to known centroids. For example, if GBIF were to flag all records within 2km of the **Monoco centroid**, we would incorrectly flag more than 27k records that [might even be in France]( http://api.gbif.org/v1/geocode/reverse?lat=43.75479&lng=7.40942). 

<img src = "/post/2019-02-21-country-centroids_files/monocoCentroid.png" style="display: block;margin-left: auto;margin-right: auto;"/>

## All provinces and countries less than 1000 sqkm
```{r, echo=FALSE, warning=FALSE,message=FALSE,fig.width=8,fig.height=4}
library(leaflet)
widget = readRDS("C:/Users/ftw712/Desktop/country centroids/rda/smallProvincesAndCentroids.rda")
widget
```

Also even if the records happen to be sitting at true centroids, they might be in a province or country so small that it is below the necessary resolution of the study. Some province centroids, such as these province centroids near **Manila** even overlap at just a 2 km buffer. 

<img src = "/post/2019-02-21-country-centroids_files/overlappingCentroids.png" style="display: block;margin-left: auto;margin-right: auto;"/>


# Less than 0.1% of GBIF records are centroids 

After excluding regions smaller than 1000 sqkm there are...

* **250 k records** within 1km of a **country** centroid
* **490 k records** within 1km of a **province or country** centroid
* **304 k records** within 2km of a **country** centroid 
* **1.3 M records** within 2km of a **province or country** centroid

So even with a large buffer (2km) there are at most **<0.1% GBIF occurrence records** sitting at or near a know centroid. This is about half as many records that have a [zero-coordiante issue](https://www.gbif.org/occurrence/search?issue=ZERO_COORDINATE), but it is still an issue probably still worth flagging on gbif.org. 

# Herbaria and museums have the most country centroids

Below are the top 15 datasets with records within 1km of a known **country** centroid. I did not included province centroids in these count totals. 

1.  [Natural History Museum Collection Specimens](https://www.gbif.org/dataset/7e380070-f762-11e1-a439-00145eb45e9a) **77 282** centroid records
2.  [The Vascular Plant Collection at the Botanische Staatssammlung München](https://www.gbif.org/dataset/7b9a08ea-f762-11e1-a439-00145eb45e9a) **34 827** centroid records
3.  [Geneva Herbarium – General Collection (G)](https://www.gbif.org/dataset/f577c9f3-ae71-4278-b6bf-512ba1dfaa21) **21 720** centroid records
4.  [Geographically tagged INSDC sequences](https://www.gbif.org/dataset/ad43e954-dd79-4986-ae34-9ccdbd8bf568) **15 469** centroid records
5.  [MAL](https://www.gbif.org/dataset/86b50d88-f762-11e1-a439-00145eb45e9a) **14 921** centroid records
6.  [Geneva Herbarium – De Candolle's Prodromus](https://www.gbif.org/dataset/63e9e22d-a649-4087-b39e-acb1651af077) **12 767** centroid records
7.  [A global database for the distributions of crop wild relatives](https://www.gbif.org/dataset/07044577-bd82-4089-9f3a-f4a9d2170b2e)  **7 057** centroid records
8.  [Field Museum of Natural History Botany Seed Plant Collection](https://www.gbif.org/dataset/90c853e6-56bd-480b-8e8f-6285c3f8d42b)  **5 877** centroid records
9.  [Botanical Museum, Copenhagen, the Phycology Herbarium](https://www.gbif.org/dataset/7f4ed916-f762-11e1-a439-00145eb45e9a)  **4 233** centroid records
10. [Field Museum of Natural History (Botany) Lichen Collection](https://www.gbif.org/dataset/aec6d756-3aa8-465e-83fd-4f6ee83e693e)  **4 216** centroid records
11. [National Museum of Natural History Luxembourg](https://www.gbif.org/dataset/962f59bc-f762-11e1-a439-00145eb45e9a)  **4 012** centroid records
12. [Norwegian Biodiversity Information Centre - Other datasets](https://www.gbif.org/dataset/492d63a8-4978-4bc7-acd8-7d0e3ac0e744)  **3 603** centroid records
13. [United States National Plant Germplasm System Collection](https://www.gbif.org/dataset/85802736-f762-11e1-a439-00145eb45e9a)  **2 995** centroid records
14. [The Fungal Collection at the Senckenberg Museum für Naturkunde Görlitz](https://www.gbif.org/dataset/7a2660bc-f762-11e1-a439-00145eb45e9a)  **2 770** centroid records
15. [Field Museum of Natural History (Botany) Bryophyte Collection](https://www.gbif.org/dataset/8a43c2b7-5976-457d-9f35-a232655efdcc)  **2 281** centroid records

Very few of these datasets fill in the [coordinate uncertainty in meters](https://terms.tdwg.org/wiki/dwc:coordinateUncertaintyInMeters) or [footprintWKT fields](https://terms.tdwg.org/wiki/dwc:footprintWKT) for the centroid points. 

# Plants make up >50% of the country centroid records

**Flowering plants** (Magnoliopsida and Liliopsida) make up more than 50% of the country centroid records on GBIF. Other groups have fewer country centroids. Mammals, reptiles, and amphibians all have less than 10k country-centroid occurrence records. Since most of the **top datasets with centroids are herbaria**, this is not surprising. 

<img src = "/post/2019-02-21-country-centroids_files/barPlotCountryCentroidsClassKey.jpg" style="display: block;margin-left: auto;margin-right: auto;"/>


# Filter out country centroids with CoordinateCleaner 

> Currently GBIF.org does not flag or filter records that fall on centroids.

Although [gbif.org](https://www.gbif.org/) does not currently filter out or flag points that might be centroids, it is still easy to do this using R. The best option is to use the R package [CoordinateCleaner](https://github.com/ropensci/CoordinateCleaner). 

```r
library(CoordinateCleaner)
library(rgbif)

# datasetkey of the Natural History Museum (London) Collection Specimens
# this dataset has around 77k centroid records of around 1M records without other geospatial issues
key <- "7e380070-f762-11e1-a439-00145eb45e9a" 

# fetch 10 000 records from NHM London Data using rgbif
NHM <- occ_data(datasetKey=key,
hasGeospatialIssue=FALSE, # remove those with other geospatial issues 
hasCoordinate=TRUE, # get only with coordinates
limit=10000)$data 

# use coordinate cleaner to clean out centroids
NHM_clean <- cc_cen(NHM,
lon = "decimalLongitude", 
lat = "decimalLatitude", 
buffer = 2000, # radius of circle around centroid to look for centroids
min_area = 1000, # min area of country or province in sqkm to check for centroids
value = "clean",
test="both")

nrow(NHM_clean) # should be around 20 records less than 10 000 

```


<!--more-->
